---
title: "Visualizing Abstracts"
author: Haley Lam
date: "2026-02-13"
output:
        html_document:
                theme: cosmo
---

The focus of this document is to compare word clouds between the three journals.

## 1) Load Data
Let's first load the data.

```{r setup, include=FALSE}
library(tidyverse)
library(tidytext)
library(wordcloud)
combined <- read_csv("Combined.csv")
```

How many rows actually have abstracts?
```{r}
sum(!is.na(combined$Abstract))
```
We have `r sum(!is.na(combined$Abstract))` available abstracts.

## 2) Tokenize Abstracts into Individual Words

Split each word in an abrstract into one row per word.
```{r}
words <- combined %>%
        filter(!is.na(Abstract)) %>%
        select(Journal, Abstract) %>%
        unnest_tokens(word, Abstract)
```
We now have `r nrow(words)` rows (i.e., words)!

Here is where I realized that the abstracts were truncated. So, these are not full abstracts, possibly because of publish & perish limitations.

## 3) Remove "Stop Words"

There's this dataset from the tidytext package that contains common words like "the", "and", "is" etc. We can use anti_join to remove those words, since they aren't of interest for our word cloud.

```{r}
words <- words %>% 
        anti_join(stop_words, by = "word")
```
There are now `r nrow(words)` number of words remaining.

Here are the top words:
```{r}
words %>%
        count(word, sort = TRUE) %>%
        head(20)
```

## 4) Make the Clouds

First, let's do JPSP:

```{r}
jpsp_freq <- words %>%
        filter(Journal == "JPSP") %>%
        count(word, sort = TRUE) %>%
        head(80)

wordcloud(
        words = jpsp_freq$word,
        freq = jpsp_freq$n,
        max.words = 80,
        scale = c(2, 0.3),
        colors = brewer.pal(8, "Dark2")
)
```

Next, let's do PSPB:

```{r}
pspb_freq <- words %>%
        filter(Journal == "PSPB") %>%
        count(word, sort = TRUE) %>%
        head(80)

wordcloud(
        words = pspb_freq$word,
        freq = pspb_freq$n,
        max.words = 80,
        scale = c(2, 0.3),
        colors = brewer.pal(8, "Dark2")
)
```

Lastly, SPPS:

```{r}
spps_freq <- words %>%
        filter(Journal == "SPPS") %>%
        count(word, sort = TRUE) %>%
        head(80)

wordcloud(
        words = spps_freq$word,
        freq = spps_freq$n,
        max.words = 80,
        scale = c(2, 0.3),
        colors = brewer.pal(8, "Dark2")
)
```


It isn't actually very clear from those word clouds. They look similar. Let's try another format:

```{r}
top_words <- words %>%
        count(Journal, word, sort = TRUE) %>%
        group_by(Journal) %>%
        slice_max(n, n = 15) %>%
        ungroup() %>%
        mutate(word_ordered = reorder_within(word, n, Journal))

ggplot(top_words, aes(x = word_ordered, y = n, fill = Journal)) +
        geom_col(show.legend = FALSE) + 
        coord_flip() + 
        facet_wrap(~ Journal, scales = "free_y") + 
        scale_x_reordered() +
        labs(x = NULL, y = "Count", title = "Most Frequent Abstract Words by Journal") +
        theme_minimal()
```

JPSP interestingly mentions "social" more than the other two journals, but also "personality"! "Relationship" consistently showed up on all three, which I assume is mostly the statistical kind of "relationship" rather than human relationships. The other words are mostly ones that are expected to appear within all kinds of research (e.g., studies/study, individuals, etc.). 