---
title: "Methodological Trends in Social/Personality Psychology (2017-2022)"
author: "Haley Lam"
date: "`r Sys.Date()`"
output: 
  html_document: 
    theme: cosmo
---
### Overview
This paper uses the existing definitions and methods validated by Lam (2025) to categorize all articles from JPSP, SPPS, and PSPB in 2017-2022. It is the "Phase 1" of 3, which aims at identifying potential for causal inference methods in psychology.


## 1. Set Envrionment and Load Data
```{r}
library(ggplot2)
library(tidyr)
library(dplyr)
library(kableExtra)
library(readr)

data <- read_csv("phase1_results.csv")
```

There are `r nrow(data)` articles in the dataset.

Now, let's select the variables that are relevant (Journal, Year, and Methodologies). Each methodology has its own column (coded 0 or 1).

```{r}
methodologies <- c("Experimental", "Causal_Inference", "Descriptive_Correlational",
                   "Theoretical", "Methodological", "Review_Meta_analysis", "Other")
data_clean <- data %>%
        select(JournalAbbr, Year, all_of(methodologies)) %>%
        rename(journal = JournalAbbr, year = Year)
```

## 2. Yearly Proportion Summaries

Now, we can reshape the data to long format and only keep the categories that have a value of 1. Since these categories are mutually exclusive (in the way they are coded, although not theoretically), there should only be one category per row.

```{r}
long_data <- data_clean %>%
  pivot_longer(
    cols = all_of(methodologies),
    names_to = "category",
    values_to = "value"
  ) %>%
  filter(value == 1)
```

Then, we can count the number of articles categorized in each methodology by journal and year (i.e., for each journal, how many articles are in each of the 6 categories within each year?)

```{r}
category_counts <- long_data %>%
  group_by(journal, year, category) %>%
  summarize(total_rows = n(), .groups = "drop")
```

Next, we can calculate those proportions, again grouping by journal and year

```{r}
category_proportions <- category_counts %>%
  group_by(journal, year) %>%
  mutate(proportion = total_rows / sum(total_rows)) %>%
  ungroup()
```

Then, we can clean up the category names (e.g., Descriptive Correlational should be Descriptive/Correlational). I made the mistake of not knowing this in my published article! 

```{r}
journal_yearly_summary <- category_proportions %>%
  mutate(
    category = gsub("_", " ", category),
    category = gsub("Descriptive Correlational", "Descriptive/Correlational", category),
    category = gsub("Review Meta analysis", "Review/Meta-analysis", category)
  ) %>%
  arrange(journal, year, category)
```

## 3. Proportion by Journal
One of the goals is to examine the proportion of articles that use a given methodology by journal.

```{r}
mean_props <- data_clean %>%
        group_by(journal) %>%
        summarize(
                Experimental = round(mean(Experimental, na.rm = TRUE), 3),
                Causal_Inference = round(mean(Causal_Inference, na.rm = TRUE), 3),
                Descriptive_Correlational = round(mean(Descriptive_Correlational, na.rm = TRUE), 3),
                Theoretical = round(mean(Theoretical, na.rm = TRUE), 3),
                N = n(),
                .groups = "drop"
        )
mean_props
```

As expected, the three social/personality journals do a lot of Experimental/Correlational work. It might be worth digging deeper into those categorized as "Causal Inference", since I have a feeling many of them aren't quite causal. 

Regarding between-journal differences, JPSP publishes the most experimental work, while SPPS does more desriptive/correlational work. This, I susepct, could be due to how experiments are well-respected in social psych? JPSP as the top journal might prefer such 'gold standard' studies, while SPPS might be more willing to take in high-quality descriptive/correlational studies.

## 4. Generate Trend Plots

Next, we can generate some plots! We can focus on the main 4 categories (as reported in Lam, 2025).

```{r}
journal_labels <- c("JPSP" = "JPSP", "PSPB" = "PSPB", "SPPS" = "SPPS")
main_categories <- c("Experimental", "Causal Inference", "Descriptive/Correlational", "Theoretical")


trends_plot <- ggplot(journal_yearly_summary %>% filter(category %in% main_categories),
                      aes(x = year, y = proportion,
                          linetype = category, shape = category,
                          color = category,
                          group = category)) +
        geom_line(linewidth = 1.2) +
        geom_point(size = 3) +
        scale_color_viridis_d(option = "D") +
        labs(x = "Year", y = "Proportion of Articles",
             linetype = "Methodology", shape = "Methodology", color = "Methodology") +
        facet_wrap(~journal, labeller = labeller(journal = journal_labels)) +
        scale_x_continuous(breaks = 2017:2022) +
        theme_bw() +
        theme(legend.position = "bottom",
              axis.text.x = element_text(angle = 45, hjust = 1))
print(trends_plot)
```

There are no obvious linear trends, though SPPS seems to be having increases in descriptive/correlational studies. This can reflect 1) the methodological conservatism found in Lam (2025) or 2) a timeframe too short to detect any meaningful trends within the field.
